Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-31 paper.tex
GPU fixes

To demonstrate the ease and efficiency with data stored in Zarr
format can be processed on a GPU, we developed two proof-of-concept
applications. 
In both cases we demonstrate on data from chromosome 22 of
1000 Genomes WGS data~\citep{byrska2022high}. This consists
of 96,514 variants and 2504 samples, converted to Zarr format
using a chunk size of 10,000 variants $\times$ 1,000 samples.
Experiments in this section were performed on a
computer with an Intel Xeon Silver 4216 processor,
Nvidia A6000 RTX GPU and NVMe SSD.

To demonstate the ease at which non-trivial, novel calculations 
can be performed on a GPU,
we first reimplemented the af-dist calculation
described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
of core operations for scientific computing.
All of the operations required to compute  af-dist are
performed on the GPU, mostly with minimal changes required
(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
We perform the central operation of genotype classification using the
CuPy JIT compiler to generate a custom CUDA kernel. 
This is very similar to the Numba compiled function used in other
experiments, within minimal changes required to take advantage of
on-GPU parallelism. In the simplest version, we iterate over the chunks
in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
a total of 7.82 MiB), decompress each to a local
NumPy array and then copy the data to GPU memory by creating
the corresponding CuPy array. For this version, the total time
in the genotype classification kernel was 25ms with a
data transfer time of 179ms (we discount decompression
time here; see next paragraph). For comparison,
our standard Numba compiled classification function required
a total of 1.46s running on a single CPU core.
Another approach in which we load the entire
\texttt{call\_genotype} array into GPU memory at once (possible
here as the $96514 \times 2504 \times 2$ array requires
460 MiB) and uses
a more sophisticated kernel in which we map GPU resources to
subsets of the array in two dimensions,
reduces transfer time to 138ms and processing time to 6ms.

>>>>

To demonstrate the ease and efficiency with which data stored in Zarr
format can be processed on a GPU, we developed two proof-of-concept
applications. 
In both cases we demonstrate on data from chromosome 22 of
1000 Genomes WGS data~\citep{byrska2022high}. This consists
of 96,514 variants and 2504 samples, converted to Zarr format
using a chunk size of 10,000 variants $\times$ 1,000 samples.
Experiments in this section were performed on a
computer with an Intel Xeon Silver 4216 processor,
Nvidia A6000 RTX GPU and NVMe SSD.

To demonstate the ease at which non-trivial, novel calculations 
can be performed on a GPU,
we first reimplemented the af-dist calculation
described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
of core operations for scientific computing.
All of the operations required to compute  af-dist are
performed on the GPU, mostly with minimal changes required
(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
We perform the central operation of genotype classification using the
CuPy JIT compiler to generate a custom CUDA kernel. 
This is very similar to the Numba compiled function used in other
experiments, with minimal changes required to take advantage of
on-GPU parallelism. In the simplest version, we iterate over the chunks
in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
a total of 7.82 MiB), decompress each to a local
NumPy array and then copy the data to GPU memory by creating
the corresponding CuPy array. For this version, the total time
in the genotype classification kernel was 25ms with a
data transfer time of 179ms (we discount decompression
time here; see next paragraph). For comparison,
our standard Numba compiled classification function required
a total of 1.46s running on a single CPU core.
Another approach in which we load the entire
\texttt{call\_genotype} array into GPU memory at once (possible
here as the $96514 \times 2504 \times 2$ array requires
460 MiB) and uses
a more sophisticated kernel in which we map GPU resources to
subsets of the array in two dimensions,
reduces transfer time to 138ms and processing time to 6ms.
